---
title: "Life Satisfacation Analysis using linear regression model"
author: "Mengying Li, Jiaxin Zhang, Chenyang Huan"
date: "Oct 19, 2020"
output: bookdown::pdf_document2
bibliography: references.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
```

## Life Satisfacation Analysis using linear regression model.

# Mengying Li, Jiaxin Zhang, Chenyang Huan 
# Oct 19, 2020

## Abstract

 With the increasing improvement of national living standard and quality, life 
 satisfaction has gradually become the focus of attention in countries, and also
 a topic of interest to our study. In this report, we mainly focus on five 
 demographic factors which are age, gender, marital status, education level and
 income levels, conduct cautious research to find out how different groups of 
 individuals in Canada have a difference in their life satisfaction. The result
 of our study found, among all the factors, marital status has the most
 significant impact on life satisfaction, and income level has the second most 
 significant impact on life satisfaction, while age, gender, and education level
 has a relatively weak impact on life satisfaction compared to other factors.

## Introduction

  In this report, our goal is to find out the factors which will affect life 
satisfaction and to what extent. In recent years, national life satisfaction 
has been a hot topic in the country. Since life satisfaction is a comprehensive
index to measure the quality of life of citizens in a country and it could also 
indirectly reflect the comprehensive level of the country, thus its research is 
of great significance to the improvement of people's quality of life. 
Our research data is from GSS, an official and authoritative database. We select 
the data of "feeling_life" from the GSS database as the research object, and 
other variables to investigate this topic together. 
With statistical knowledge to conduct a cautious research survey, we aim to 
investigate which factor of these five factors will have a greater or lesser 
impact on people's life satisfaction. We build a linear regression model with 
the sorted data. There are five independent variables, namely age, gender, 
marital status, income level and education level. The dependent variable is life 
satisfaction, which is the topic we want to investigate. The conclusion of this 
report is marital status is the most significant factor affecting life 
satisfaction. Age and gender have an approximately statistically significant 
but smaller effect on life satisfaction to the extent. Compared with the first 
three factors, income level has a less statistically significant but relatively 
large impact on life satisfaction. Education level on life satisfaction has the 
weakest effect among five variables, most categories in education have an 
insignificant impact on life satisfaction except for the category less than high
school. 
This model can help us to analyze the relationship between life satisfaction and 
these factors and predict the level of life satisfaction under given conditions. 
Moreover, through the linear regression model, we can also make specific 
solutions to the improvement of life satisfaction under different conditions.



## Data

The 2017 General Social Survey which is available on chass in utoronto library
  was conducted from February 2$^{nd}$ to 
  November 30$^h$. The objectives are collecting social trends data to track 
  changes in both living conditions and well-being of Canadians. In addition, 
providing information on current or with emerging interest policies. As this 
dataset is the latest version available, it would be appropirate to make 
analysis.The focus 
of this report is what factors would affect Canadians' life satisfaction, so 
feelings_life
is the response variable while age, gender, region, province, income_respondent, 
education, marital_status, children_in_household, owe_rent will be investigated
to find the relationship between the response variable and them. 
  Core content and classification variables were included in order to achieve 
  the objectives. The core content was created to measure social changes related
  to life and provide information to report on welfare and other specific 
  policies. 
  The survey was distributed through out phone calls. The target population is 
  people live in Canada who aged 15 and above while residents in Yukon Northwest
  Territories, Nunavut and full-time residents of institutions were excluded.
  The frame was made up of the list of telephone numbers in use that were 
  available on Statistics Canada and the list of all dwellings on Address 
  Register. The General Social Survey used stratified sampling, every record 
in the frame was allocated to a stratum within its province, then a simple 
random sampling without replacement was performed within those stratum. 
Trade-offs might be some areas were over-represented while others were 
under-represented and some provinces were excluded. 
Among the phone number list, 91.8% of them reaches eligible households that 
have at least one person who was aged 15 and above. The sample population was 
20,602. 52.4% of the phone call survey was responded, since "cell phone only" 
households were added to the frame and these population was harder to reach.  
There were three stages of adjustments for non-responses based on amount of 
auxiliary information of the phone number available in Statistics Canada and 
weight for responding telephone numbers would be adjusted accordingly.
  The questionnaire was quite comprehensive and questions were all related to 
  target population's well-being so that appropriate adjustment of social 
  policies could be made based on the data collected. Some questions and options
  in the questionnaire might be misleading and caused some non-response. For 
  instance, the question asked for respondent's macro-region's place of birth 
  has lots of NAs for those who were born in Canada. The answer categories 
  includes "Americas",""Europe" and other continents, people might consider 
  "Americas" as the United States.



## Model
This report is interested in one topic, namely: the satisfaction with the life 
of people of different ages, genders, regions, income levels, and family status. 
To study this topic, we selected the following variables from the data set: 
feelings_life, age, gender, region, province, income_respondent, education, 
marital_status, children_in_household, owe_rent. We write code in 
Rstudio which is statistical software that able to build and summarize models. 
To investigate life satisfaction, we choose the variable feelings_life as the 
response variable (Y) of our model, which is a quantitative variable from 0 to 
10 
(from "very dissatisfied" to "very satisfied" respectively). This variable can 
be treated as an 11-category variable. However, since we use this variable as a 
response variable, if we treated it as a category variable, we should use a 
logical regression model which is a statistical method that enables us to 
summarize and study relationships between the category response variable and 
multiple predictor variables. Logistical regression only accepts the response 
variable with binary category while we aim for a specific numeric result on our 
model other than binary categories such as satisfied and dissatisfied. 
Considering that binary life satisfaction cannot provide a more detailed result, 
we choose to treat life satisfaction as a quantitative variable. Therefore, we 
use the multivariable linear regression in building the model, which allows us 
to summarize and study relationships between the quantitative response variable 
and multiple predictor variables. 
The first predictor is the age which is a numeric variable. Choosing numeric 
ages 
instead of age-groups is because numeric variable is more specific. Moreover, 
one-parameter rather than several parameters (for different categories) can save 
unnecessary complexity (using age-groups requires us to divide the age into many 
groups which result in many more parameters in our model, thus may lead to 
overfitting, which means that the model is too complex and difficult to 
interpret and understand). 
The remaining eight variables are all categorical because these variables are 
variables composed of different category groups that enable us to investigate 
the different relationships with our response variable feelings_life. For 
example, 
sex is in two categories: male and female. We want to see whether these 
different 
types of sex have different relationships with life satisfaction. On the other 
hand, personal income and the number of children in a household can be treated 
as numeric variables, but we collect the data in categorical groups. Personal 
income is categorical instead of numeric because we want to investigate the 
difference in life satisfaction among people of different income levels other 
than a specific number. The number of children in a household only range 
from 0 to 2, and the majority is zero, so we keep this variable categorical
in the hope of higher clarity. Since provinces and regions are remarkably 
similar variables, we chose to use the region variable to avoid overfitting.
While other variables are self-explanatory, the own_rent represents the status
of whether the respondent is living in an owning or renting property.
The dataset used stratified sampling in different provinces in Canada, which 
requires the population size of each province in model building. According to 
Statistics Canada, the population size by 2017 season 4 for each province is 
as following: Quebec: 8336460; Manitoba: 1340565; Ontario: 14150783; 
Alberta: 4258195; Nova Scotia: 953683; British Columbia: 4956550; 
Saskatchewan: 1154755; Prince Edward Island: 151376; New Brunswick: 767935; 
Newfoundland and Labrador: 528288. 

We continue our analysis by building a linear regression model to predict the 
variable feelings_life using the above-mentioned predictor variables 
(except for province). Table 1 in the result section shows the summary of this 
model. P-values for the variable region, own_rent, chiledren_in_household are 
insignificant (large value), which indicates a large possibility of validation 
of our null hypothesis (no difference in the response variable). Hence, we 
decide 
to drop these three variables out of our model. By using age, sex, marital 
status, 
education, and income variables, we build a linear regression with better 
performance than the previous alternate model. 
The formula of this model is as follow:

$$
Feelingslife = \beta_0 + \beta_1 x_{age} + \beta2x_{sex,Male} + \beta3x_{maritalstatus,ComLaw} + \beta4x_{maritalstatus,Married} +\\beta5x_{maritalstatus,Separated} +\beta6x_{maritalstatus,Single} +\beta7x_{maritalstatus,Widowed} +\\beta8x_{education,College} +\beta9x_{education,Highschool} +\beta10x_{education,LessHign} +\\beta11x_{education,Trade} +\beta12x_{education,UnivbelowBach} +\ \beta13x_{education,UnivaboveBach} + \\beta14x_{income,A} +\beta15x_{income,B} + \beta16x_{income,C} + \beta17x_{income,D} + \beta18x_{income,E} + \epsilon_i
$$
when income is A refers to the income level of higher than $12500, income B 
refers to the income level of $25000-$49999, income C refers to the income level 
of $50000-$74999, income D refers to $75000-$99999 and income E refers to income 
level of less than $25000 (when income level is $10000-12499, it has no effect 
on our response variable, this applies to all the category not showing in the 
the formula for all categorical variables such as female). Other variables are 
self-explanatory. The beta0 is the intercept value. It means the value of 
feeings_life when all other variable equals to zero, which does not apply in 
our model.

The formula tells us, the value of feelings_life (life satisfaction) equals to 
the sum of the parameters of variables with the respective category. For 
example, for a single female with an age of 30, education level of college and 
income level of$50000 per year, we expect that her value of feelings_life equals
to the intercept(beta0), plus parameter beta1 times 30(her age), plus parameter 
beta6, plus parameter beta8, plus parameter beta16 (with small errors). 



Table 2 in the result section shows the summary of our model. The P values are 
significant (extremely small) for every predictor variable (some category of the 
variable has insignificant P-value), which indicates a large possibility of 
validation of our alternate hypothesis (statistical difference in the response 
variable). That is, variables in the model are statistically significant to 
predict the value of our response variable, feelings of life. 
However, the low R-squared value shows that only a small portion of data is 
explained by our estimates. Another caveat is that not all of the model 
assumption is satisfied such as Linearity of this model are not well satisfied. 
Besides, this model only applies to Canadians given our dataset. 

## Results

```{r load data, include=FALSE}
library(survey)

# load the 2017 gss dataset
setwd('/Users/chenyanghuan/Desktop')
DATA <- read.csv("gss.csv")

#If any categorical data are not factor type, convert it to factor 
DATA$sex<-as.factor(DATA$sex)
DATA$region<-as.factor(DATA$region)
DATA$marital_status<-as.factor(DATA$marital_status)
DATA$education<-as.factor(DATA$education)
DATA$own_rent<-as.factor(DATA$own_rent)
DATA$self_rated_health<-as.factor(DATA$self_rated_health)
DATA$self_rated_mental_health<-as.factor(DATA$self_rated_mental_health)
DATA$income_respondent<-as.factor(DATA$income_respondent)
DATA$children_in_household<-as.factor(DATA$children_in_household)
DATA$province<-as.factor(DATA$province)


```

```{r selected interested varoables, include=FALSE}

y <- DATA %>%
  select(feelings_life)

x1 <- DATA %>% select(age)
x2 <- DATA %>% select(sex)
x3 <- DATA %>% select(region)
x4 <- DATA %>% select(marital_status)
x5 <- DATA %>% select(education)
x6 <- DATA %>% select(own_rent)
x7 <- DATA %>% select(income_respondent)
x8 <- DATA %>% select(children_in_household)
x9 <- DATA %>% select(province)

indata <- data.frame(c(y,x1,x2,x3,x4,x5,x6,x7,x8,x9))
```

```{r drop NA, include=FALSE}
#If there are any NAs, drop it.

indata<-na.omit(indata)
```

Figure \@ref(fig:fig1) shows that most of the observations having a value of 
greater than 7 on feeling_life. Plus, the median is 8 while the mean is 8.094. 
This suggests a high level of life satisfaction in our dataset. Figure 1 shows 
11 categories as we mentioned in the model section. However, simply dividing the 
variable into a binary category would be less interpretable. If we simply use 
the median value as the dividing line, then values below 8 will be attributed to
dissatisfaction, which will make our results inconsistent with common sense. 
If we adopt a reasonable value such as 6 as the dividing line, then the number 
of people in the group below 6 will be too small to affect the accuracy of our 
results. Hence, again, we choose the linear regression to build models.

```{r fig1, fig.cap="feelings of life"}
indata %>% ggplot(aes(feelings_life)) + geom_histogram()
indata %>% summarise(AVG = mean(feelings_life, na.rm=T),
                                                Med = median(feelings_life, na.rm = T))
```

Figure \@ref(fig:fig2) shows the relationship between feelings_life and each 
predictor. The plots show that there are observable differences in the value of 
feelings_life between the different categories in variables: marital_status, 
income_respondent, own_rent. While other variables do not show differences, 
we still need to see our model result to make an interpretation.

```{r fig2, fig.cap="Plots of relationship between response variable and each predictor "}
a <- indata %>% ggplot(aes(y=feelings_life, x=province)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

b <- indata %>% ggplot(aes(y=feelings_life , x= region)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

c <- indata %>% ggplot(aes(y=feelings_life , x= age)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

d <- indata %>% ggplot(aes(y=feelings_life, x=sex)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

e <- indata %>% ggplot(aes(y=feelings_life , x=income_respondent)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

f <- indata %>% ggplot(aes(y=feelings_life, x= marital_status)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

g <- indata %>% ggplot(aes(y=feelings_life , x= education)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

h <- indata %>% ggplot(aes(y=feelings_life , x= own_rent)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

i <- indata %>% ggplot(aes(y=feelings_life , x= children_in_household)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

library(gridExtra)
grid.arrange(a,b,c,d)
grid.arrange(e,f)
grid.arrange(h,i)
h
g
```

Table \@ref(tab:table1) has the results of the alternate linear regression 
model. we can see that all of the P-value for categories in the 
children_in_household
variable is larger than 0.05 which indicates they are statistically 
insignificant 
for predicting our response variable. Also, most P-values for categories in the 
region and own_rent variables are insignificant either.Moreover, as previously
seen in figure 2, different categories in these 3 variables have minimal
differences in the mean of our response variable. Hence, it is reasonable to 
drop these 3 variables in our final model. 

```{r table1}
indata$popsize<-ifelse(indata$province=="Quebec", 8336460,ifelse(indata$province=="Manitoba", 1340565,ifelse(indata$province=="Ontario",14150783,ifelse(indata$province=="Alberta",4258195,ifelse(indata$province=="Nova Scotia",953683,ifelse(indata$province=="British Columbia",4956550,ifelse(indata$province=="Saskatchewan",1154755,ifelse(indata$province=="Prince Edward Island",151376,ifelse(indata$province=="New Brunswick",767935,ifelse(indata$province=="Newfoundland and Labrador", 528288, 0))))))))))

svydesign_strs<-svydesign(id=~1,strata=~province, data=indata, fpc=~popsize)
mod <- svyglm(feelings_life ~ age+sex+region+own_rent+marital_status+education+income_respondent+children_in_household, svydesign_strs, family="gaussian")

kable(summary.lm(mod)$coefficients, caption="Summary of the selected model", digits=3)

```

```{r table2}
mod2 <- svyglm(feelings_life ~ age+sex+marital_status+education+income_respondent, svydesign_strs, family="gaussian")

kable(summary.lm(mod2)$coefficients, caption="Summary of the selected model", digits=3)
```


Table \@ref(tab:table2) has the results of our final selected model. 
First, we can see that all of the P-value for categories in the marital_status 
variable is extremely small than 0.05, which indicates it is statistically 
significant for predicting our response variable. Also, all the significant 
P-values suggest that the most important feature in our model is marital status.
The result shows that the marital status of common-law and married shows a 
significant large positive effect on the feelings of life compared to other 
categories. Plus, the marital status of single and widowed has a significant 
small positive effect on the feelings of life, while the separated marital 
status has a significant negative effect on the feelings of life. 
Other significant variables include age (which indicates a minimal positive 
relationship with feelings of life), sex male (indicates a negative relationship
with feelings of life compared with sex female), education of less than high 
school diploma (indicates a small positive relationship with feelings of life
compared with all other education categories), the income level of higher than 
$125000 (indicates a small positive relationship with feelings of life), the 
income level of $25000-$49999 (indicates a small negative relationship with 
feelings of life), and income level of less than $25000 (indicates a relatively
large negative relationship with feelings of life compared with other income 
levels).

Figure \@ref(tab:figure3) has the results of the model check. The figure shows 
that multicollinearity, independence, and constant variance assumptions are fine
in our model. However, normality and linearity assumptions are not well 
satisfied. With the large sample size (more than 20000), normality can be
ignored.The dissatisfaction of linearity assumption may suggest transformation 
on the model for better performance, which indicates a weakness of this model.

```{r figure3}
#assumption check/ model check
par(mfrow=c(3,3))
qqnorm(rstandard(mod2))
qqline(rstandard(mod2))
plot(indata$feelings_life ~ mod2$fitted.values, xlab="Fitted", ylab="Feelings of Life")
abline(a = 0, b = 1, lty=2)
lines(lowess(mod2$fitted.values, indata$feelings_life), col="red")

plot(rstandard(mod2) ~ fitted(mod2), xlab="Fitted", ylab="Residuals")
plot(rstandard(mod2) ~ indata$age, xlab="Age", ylab="Residuals")
plot(rstandard(mod2) ~ indata$sex, xlab="Sex", ylab="Residuals")
plot(rstandard(mod2) ~ indata$marital_status, xlab="Matital Status", ylab="Residuals")
plot(rstandard(mod2) ~ indata$education, xlab="Education", ylab="Residuals")
plot(rstandard(mod2) ~ indata$income_respondent, xlab="Income", ylab="Residuals")
```


## Discussion

From the final selected model we could see:
First, we discuss the impact of marital_status on life satisfaction. obviously, 
all of the p-value for categories in marital_status variable is extremely small
than 0.05 which indicates the significant for predicting life satisfaction. In 
detail, only the marital status of separated has a negative effect on life 
satisfaction, other marital status all have the positive effects on life 
satisfaction. Among them, the marital status of married has the most significant 
positive effect on life satisfaction than other marital status, which could 
be reflected in the real life that married people always tend to feel the 
happiest than other people.
Second, we discuss the impact of age on life satisfaction. Since the variable of
age is a numerical variable, we could get the result from this model that, when 
other conditions remain unchanged, when the age increases by one year, the life 
satisfaction will increase by 0.0027 units. It can be reflected in real life, 
that is, life happiness will increase with the increasing of age.
Third, we discuss the impact of sex on life satisfaction. The p-value for the 
variable of age is extremely small, so we could tell that it is significant for 
predicting life satisfaction. The negative number shows that male have a 
negative relationship on life satisfaction compared with female. In the real 
life, we could get a general conclusion that Women are more likely to feel 
happy than men.
Next, we discuss the impact of education level on life satisfaction. By 
observing the p-value of those categories in education variable, we can get that 
except the education of less than high school diploma or its equivalent has a 
small positive effect effect on life satisfaction, other categories of education 
variable have no relationship or impact with the life satisfaction since those 
p-values are much bigger than 0.05. Hence, we consider that the variable of 
education level have a very insignificant linear relationship with life 
satisfaction, or even no relationship. This model can reflect in real life that 
people will not be affected about their happiness because of their education 
level.
In the end, we discuss the impact of income on life satisfaction. By looking at 
the p-value, we can see that income level of $50,000-$74,999 and $75,000-$99,999
may not have relationship with life satisfaction since their p-values are much 
bigger than 0.05. However, other income levels have the significant effect on 
life satisfaction since their p-value are extremely small. In those levels, only
income of $125,000 and more indicates a small positive relationship with life 
satisfaction, income of $25,000-$49,999 and income of less than $25,000 both 
show negative relationships with life satisfaction while income of less than 
$25,000 has more huge negative effect compared to another income level. From
this result, we could tell that income will effect the life satisfaction in the 
real life. Only when income more than $125,000, people will satisfied their 
life, Or they will not satisfied. And when the income level reaches a relatively
low level like less than $25,000, the signifiant negative effect on people's 
life satisfaction will occur immediately.

# Weaknesses

Weakness for the data is that residents from Yukon Northwest Territories, 
Nunavut and full-time residents of institutions and people aged below 15 are 
excluded, so the result might not apply to all Canadians.In addition, the 
response rate is relatively low comparing to previous years’ data collection due 
the updated population frame. Since linearity of this model is not satisfied, 
the regression algorithm might not mathematically capture the trend and lead to 
an inefficient model.

# Next Steps

Health condition could also be included in the questionnaire, as high levels
of life satisfaction often associates with body wellness. The overall  health 
condition both physically and mentally would affect feelings of life 
dramatically. In the future, there could be a subsequent survey for people
who have feelings of life <=3 to track their mental wellness and give out 
support when necessary. 

## References

\@ref(tab:table1)

\@ref(fig:fig1)

[@citeR]


